Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2017-12-07 paper.tex
cutting of text, algorithm updated

\author{Vojt\v ech Von\' asek$^{1}$,
    Barbora Kozl\'\i kov\'a$^{2}$,
    Adam Jur\v{c}\'\i k$^{2}$,
    Ond\v{r}ej V\'{a}vra$^{3}$,
    Martin Saska$^{1}$
\thanks{$^{1}$ Faculty of Electrical Engineering,  Czech Technical University in Prague, 
Technick\'a 2, 166 27 Prague, Czech Republic,
{\tt vonasek@labe.felk.cvut.cz}}
\thanks{$^{2}$        
Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic}
\thanks{$^{3}$        
Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic}
%The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
%Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
%}%
%}
}

@@

\begin{figure}[t]
\centering
{\footnotesize
\renewcommand{\arraystretch}{-0.5}
\renewcommand{\tabcolsep}{-1pt}
\begin{tabular}{ccc}
\includegraphics[width=0.15\textwidth]{fig/motiv1} &
\includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
\includegraphics[width=0.16\textwidth]{fig/motiv3}  \\
Protein 1CQW & Active site & Detected tunnels \\ %& Example of a  \\
             &            & (orange)         \\  %& trajectory
\multicolumn{3}{c}{%
\includegraphics[width=0.16\textwidth]{fig/renderDCP}  \hskip 15pt
\includegraphics[width=0.14\textwidth]{fig/render37t}} \\ 
\end{tabular}
}
\caption{\label{fig::motiv}
    Tunnels in the DhaA haloalkane dehalogenase protein (PDB ID 1CQW) with possible trajectories of 1-Chlorpropan ligand (bottom left) and 1,2-Dichloroethane (bottom right).
}
\end{figure}

\textcolor{blue}{To provide the biochemists with a solution which detects tunnels also in these frames and understand the ligand passage, it is necessary to take into account not only the ligand shape but also its conformation changes when the ligand moves.}
Such a ligand, changing the mutual position of its atoms within a molecular dynamics, is denoted as a flexible ligand.
We can formulate this task as a motion planning problem.
The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.

The idea of the sampling-based motion planning is to randomly sample the configuration space of the robot and to connect the collision-free samples
into a graph structure (roadmap).
A path in the roadmap then corresponds to a motion in the workspace.
%The random samples are classified as free or non-free using black-box collision detection which allows the sampling-based methods to find paths for robots of various shapes. 
%The sampling-based planners can cope with robots with many degrees of freedom (DOF).
Two most commonly used sampling-based techniques are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT}.
The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of an arbitrary shape.
Therefore, they have been used for motion planning of various robotic systems.
Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
%Sampling-based motion planners has been intensively studied in robotics, but they have been applied also in other research fields.
%In bio-chemistry, the planners have been used to study
to study loop motions~\cite{cortes2004geometric}
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
and for protein folding~\cite{raveh2009rapid,novinskaya2015improving,amato2002using}.
%or for tunnel detection~\cite{vonasek2017tunnel}.

In this paper, we propose a modification to RRT to address the problem of planning a non-spherical flexible ligand passage through a given protein.
We localize the generation of the random samples, i.e., instead of sampling the whole configuration space we focus only on the space surrounding a potential tunnel detected by the Voronoi-based method (using a smaller ligand probe in order to avoid omitting possibly feasible tunnel candidates).
The potential tunnel is represented by a set of consecutive spheres surrounding the tunnel centerline.
This optimization allows us to find trajectories along the given tunnel, which is faster than searching for any trajectory in the whole protein.
To enable the movement of the ligand in the narrow parts of the tunnel, the free-space can be dilated by shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}. 
The ligand flexibility is modeled using a fixed set of predefined conformations which the ligand can occupy.

@@

Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or they utilize a grid-based computation~\cite{sehnal2013mole}.
%The tunnels are pathways computed for a single atom (spherical probe) using Voronio-diagrams~\cite{yaffe2008,caver3} or 
%grid-based approaches~\cite{sehnal2013mole,petrek2006caver}.
The resulting tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and a list of surrounding residues, whose physico-chemical properties are crucial for assessing the interaction possibilities between the protein and ligand.
More details can be found in the recently published state-of-the-art report about analysis and visualization of biomolecular cavities~\cite{Krone_2016}.
The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand might traverse the tunnel.
Moreover, the ligand flexibility cannot be principally considered by these approaches.
Simulations based on molecular dynamics (MD) can directly evaluate the migration of a ligand into a protein, but they are significantly more computationally demanding~\cite{kingsley2014including}. 
Moreover, MD simulations have to be prepared for a specific ligand/protein pair, which is not practical for testing a large number of ligands.
The geometric-based methods are therefore still preferable due to their speed.
%In order to determine if the ligand can pass the tunnel using the geometric-based methods, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.

@@

\section{Proposed Method}
To address the problems of the existing solutions for computing the non-spherical flexible ligand trajectories through protein tunnels, we propose a modification of the RRT planner. 
Our approach takes into account the shape and flexibility of the ligand and localizes the searched configuration space using the information about tunnels detected by the Voronoi-based approach.
The input is therefore formed by a set of tunnels computed using Voronoi diagrams and the task is to find trajectories for a ligand along these tunnels
leading from outside the protein to the active site.
The tunnels are detected in a sequence time steps capturing protein dynamic behavior computed using MD simulations.

In contrast to~\cite{cortes2010simulating}, which searches for any exit path from the protein, we localize the searching and compute the trajectories only along each tunnel separately.
This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
Computing the trajectories along a single tunnel brings several advantages from the motion planning point of view as well.
It decreases the volume of the configuration space to be searched and increases the relative volume of the narrow passages.
To further cope with the narrow passage problem, the random samples are not generated in the whole configuration space, but only (mainly) around
the tunnel in a user-defined distance.
Therefore, the tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
A protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ is their 3D position, and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
Tunnels serving as an input for our proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.

%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
%knowledge-based methods have also been used.[17]
%13 = https://link.springer.com/article/10.1007/BF00123666

%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
The flexibility of the ligand is modeled using the set $\L$ of its conformations.
We assume that the library of all possible conformations in a considered resolution is available (e.g., ~\cite{dunbrack}) and we later show how to select a subset of conformations suitable for the motion planning.
%The conformations $\L$ are used from a library (e.g.,~\cite{dunbrack}), or they can be prepared considering the potential energy.

To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
In our solution we use a discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (its geometric center), the rotation around $x$, $y$, and $z$ axes,
the index of the conformation $l\in \L$, and the scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various spatial conformations.
The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
The initial configurations are searched around the beginning of the tunnel.
To find a new initial configuration, a random sample $q$ is generated around the first sphere of the tunnel 
$c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
If the sample $q$ is collision-free, it can be considered as a new starting configuration and it is added to $\QI$.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.

@@

The second proposed modification is the expansion procedure (Alg.~\ref{alg::expand}), which expands the tree from $\qnear$ towards $\qrand$.
The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line in the 
distance $\varepsilon$ from $\qnear$.
However, in the case of ligands, this would mostly lead to a collision.

The expansion procedure attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
and tested for collisions.
The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
part $(x,y,z)$ is generated around $\qnear$.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales.
This forces the planner to automatically use smaller scales in the difficult (narrow) parts of the tunnels, while preferring
large scales if possible.

@@

In the expansion step, each conformation from $\L$ is tested in order to approach $\qrand$.
By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
On the other hand, using a large number of conformations also increases the time complexity of the expansion step.
Selection of a suitable set of conformations $\L$ is therefore crucial.
Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are closer to each other, but 
the atoms can also form a more elongated shape (Fig.~\ref{fig::m040c}).
From the motion planning point of view, both types of conformations are useful.
The longer conformations can better fit in a narrow passage, while the packed conformations can be better rotated at a given position.
In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
packed conformations can be used.
\vspace{-3mm}

\begin{figure}[hb]
\centering
\includegraphics[width=0.14\textwidth]{fig/m040-conf1} \hskip 25pt
\includegraphics[width=0.11\textwidth]{fig/m040-conf2}
\caption{\label{fig::m040c}
 The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
}
\end{figure}

@@

The proposed approach was used to analyze the ligand traversability towards the active site using the main tunnel in the Haloalkane dehalogenase protein (PDB ID 1CQW).
The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
The molecular dynamics of the protein was computed using the AMBER tool~\cite{amber}.
In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
In most of the frames, two tunnels were detected.
The average bottleneck sizes of these tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the main and the side tunnel, respectively.
The tunnel traversability was analyzed in 100 time steps (frames) of MD simulation.

@@

Two motion planners were utilized: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
The \RA\ method was run with the following parameters:
$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
Both methods were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners as it resulted in the highest success rates.

@@

\begin{table}[ht]
\centering
\caption{\label{tab::one}Traversability for one atom.}
\renewcommand{\tabcolsep}{4.3pt}
{\small
\renewcommand{\arraystretch}{0.7}
\input graphProbe.tex
}
\end{table}

@@

\begin{table}[ht]
\centering
\caption{\label{tab::selection}
    The influence of the conformation selection method to the traversability rates in the first tunnel.
    The number after~`$/$' denotes the number of atoms.
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\small
\renewcommand{\arraystretch}{0.7}
\input graphSel.tex
}
\end{table}

@@

Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane), and
the task was to compute the trajectories for these ligands in a given molecular dynamics simulation.

@@

\begin{table}[bt]
\centering
\caption{\label{tab::main}
    Traversability of the tunnels for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\small
\renewcommand{\arraystretch}{0.7}
\input graph.tex
}
\end{table}

@@

{\def\a{270}
\begin{figure}[hb]
\centering
%\includegraphics[width=0.17\textwidth,angle=\a]{fig/m037tmins-03tunnel-1rrt_path_time-2}
%\includegraphics[width=0.17\textwidth,angle=\a]{fig/m037tmins-03tunnel-1trajDistToGoal-2}
\includegraphics[width=0.18\textwidth,angle=\a]{fig/crop1}
\includegraphics[width=0.18\textwidth,angle=\a]{fig/crop2}
\caption{\label{fig::comparison}
    Runtimes and the distance to the active site for the m037 ligand at scale 0.6.
    The values are computed from all 200 trials (20 initial configurations $\times$ 10 trials).
    Only each third frame is shown in the graph.
}
\end{figure}
}

>>>>

\author{Vojt\v ech Von\' asek$^{1}$,
    Barbora Kozl\'\i kov\'a$^{2}$,
    Adam Jur\v{c}\'\i k$^{2}$,
    Ond\v{r}ej V\'{a}vra$^{3}$,
    Martin Saska$^{1}$
\thanks{$^{1}$ Faculty of Electrical Engineering,  Czech Technical University in Prague, 
Technick\'a 2, 166 27 Prague, Czech Republic,
{\tt vonasek@labe.felk.cvut.cz} $^{2}$        
Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic 
$^{3}$        
Loschmidt Laboratories, Department of Experimental Biology RECETOX,Faculty of Science, Masaryk University
    Kamenice 5, 625 00 Brno, Czech Republic}
    }
%The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
%Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
%}%
%}
%}

@@

\begin{figure}[t]
\centering
{\footnotesize
\renewcommand{\arraystretch}{-0.5}
\renewcommand{\tabcolsep}{-1pt}
\begin{tabular}{ccc}
\includegraphics[width=0.14\textwidth]{fig/motiv1} &
\includegraphics[width=0.15\textwidth]{fig/motiv2lab} &
\includegraphics[width=0.14\textwidth]{fig/motiv3}  \\
Protein 1CQW & Active site & Detected tunnels \\ %& Example of a  \\
             &            & (orange)         \\  %& trajectory
\multicolumn{3}{c}{%
\includegraphics[width=0.16\textwidth]{fig/renderDCP}  \hskip 15pt
\includegraphics[width=0.14\textwidth]{fig/render37t}} \\ 
\end{tabular}
}
\caption{\label{fig::motiv}
    \small
    Tunnels in the DhaA haloalkane dehalogenase protein (PDB ID 1CQW) with possible trajectories of 1-Chlorpropan ligand (bottom left) and 1,2-Dichloroethane (bottom right).
}
\end{figure}

\textcolor{blue}{To provide the biochemists with a solution which detects tunnels also in these frames and understand the ligand passage, it is necessary to take into account not only the ligand shape but also its conformation changes when the ligand moves.}
Such a ligand, changing the mutual position of its atoms within a molecular dynamics, is denoted as a flexible ligand.
We can formulate this task as a motion planning problem and employ sampling-based methods to solve it.
%The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.

Sampling-based planners randomly sample the configuration space of the robot and to connect the collision-free samples
into a graph structure (roadmap).
A path in the roadmap then corresponds to a motion in the workspace.
The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of an arbitrary shape.
Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
to study loop motions~\cite{cortes2004geometric}
and for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving
%The random samples are classified as free or non-free using black-box collision detection which allows the sampling-based methods to find paths for robots of various shapes. 
%The sampling-based planners can cope with robots with many degrees of freedom (DOF).
%Two most commonly used sampling-based techniques are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT}.
%Therefore, they have been used for motion planning of various robotic systems.
%Sampling-based motion planners has been intensively studied in robotics, but they have been applied also in other research fields.
%In bio-chemistry, the planners have been used to study
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%or for tunnel detection~\cite{vonasek2017tunnel}.

In this paper, we propose a modification to Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner  to address the problem of planning a non-spherical flexible ligand passage through a given protein.
We localize the generation of the random samples, i.e., instead of sampling the whole configuration space we focus only on the space surrounding a potential tunnel detected by the Voronoi-based method (using a smaller ligand probe in order to avoid omitting possibly feasible tunnel candidates).
The potential tunnel is represented by a set of consecutive spheres surrounding the tunnel centerline.
This optimization allows us to find trajectories along the given tunnel, which is faster than searching for any trajectory in the whole protein.
To enable the movement of the ligand in the narrow parts of the tunnel, the free-space can be dilated by shrinking the atom radii, similarly, e.g., to~\cite{cortes2005path,hsu06multilevel}. 
The ligand flexibility is modeled using a fixed set of predefined conformations which the ligand can occupy.

@@

Most of the available tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or they utilize a grid-based computation~\cite{sehnal2013mole}.
%The tunnels are pathways computed for a single atom (spherical probe) using Voronio-diagrams~\cite{yaffe2008,caver3} or 
%grid-based approaches~\cite{sehnal2013mole,petrek2006caver}.
The tunnels are represented as a sequence of spheres and characterized by their bottleneck, length, curvature, and a list of surrounding residues, whose physico-chemical properties are crucial for assessing the interaction possibilities between the protein and ligand.
More details can be found in the recently published state-of-the-art report about analysis and visualization of biomolecular cavities~\cite{Krone_2016}.
The main disadvantage of both grid-based and VD-based methods is that the shape of the ligand is not taken into account during the tunnel detection and it is therefore not easy to estimate if (and how) a non-spherical ligand might traverse the tunnel.
Moreover, the ligand flexibility cannot be principally considered by these approaches.
Simulations based on molecular dynamics (MD) can directly evaluate the migration of a ligand into a protein, but they are significantly more computationally demanding~\cite{kingsley2014including}. 
Moreover, MD simulations have to be prepared for a specific ligand/protein pair, which is not practical for testing a large number of ligands.
The geometric-based methods are therefore still preferable due to their speed.
%In order to determine if the ligand can pass the tunnel using the geometric-based methods, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.

@@

\section{Proposed Method}
%To address the problems of the existing solutions for computing the non-spherical flexible ligand trajectories through protein tunnels, we propose a modification of the RRT planner. 
We propose a modification of RRT to compute trajectories for non-spherical flexible ligands through protein tunnels.
%Our approach takes into account the shape and flexibility of the ligand and localizes the search around given tunnels. configuration space to 
%using the information about tunnels detected by the Voronoi-based approach.
The input is formed by a set of tunnels computed using Voronoi diagrams and the task is to find trajectories for a ligand along these tunnels
leading from outside the protein to the active site.
The tunnels are detected in a sequence time steps capturing protein dynamic behavior computed using MD simulations.

In contrast to~\cite{cortes2010simulating}, which searches for any exit path from the protein, we localize the searching and compute the trajectories only along each tunnel separately which allows biochemists to analyze selected tunnels.
%This is motivated by the practical needs of biochemists that are interested in the actual traversability the ligand through a selected tunnel.
Computing the trajectories along a single tunnel brings several advantages from the motion planning point of view as well.
It decreases the volume of the configuration space to be searched and increases the relative volume of the narrow passages.
The random samples are not generated in the whole configuration space, but only around the tunnel in a user-defined distance which helps to cope with the narrow passage problem.
%The tunnels serve as a navigation path for the configuration tree~\cite{vonasek2009rrt}.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere representing an atom is given by its van der Waals radius.
A protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ is their 3D position, and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
Tunnels serving as an input for our proposed method can be detected using some of the existing computational tools, such as CAVER 3.0~\cite{caver3}.
%Conformations of the ligand may be generated in the absence of the receptor and subsequently docked[13] 
%or conformations may be generated on-the-fly in the presence of the receptor binding cavity,[14] 
%or with full rotational flexibility of every dihedral angle using fragment based docking.[15] 
%Force field energy evaluation are most often used to select energetically reasonable conformations,[16] but 
%knowledge-based methods have also been used.[17]
%13 = https://link.springer.com/article/10.1007/BF00123666
%The proteins are dense structures and the tunnels are typically narrow.
%Depending on the protein, the tunnel bottlenecks can be even less than $1$~\AA, which is too narrow for ligands with more than 2 atoms.
The ligand is modeled using the set $\L$ of its conformations.
The conformations are prepared using the Rosetta tool and ranked according their potential energy. Only conformations
with energy lower than a user-defined threshold are used.
%The flexibility of the ligand is modeled using the set $\L$ of its conformations.
%We assume that the library of all possible conformations in a considered resolution is available
%(e.g., ~\cite{dunbrack}) 
%and we later show how to select a subset of conformations suitable for the motion planning.
To enable the motion of ligands in the narrow tunnels, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
In our solution we use a discrete set of scales, i.e., $s \in \S=\{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale, and $\sdelta$ is the minimal difference between two scales.

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (its geometric center), the rotation around $x$, $y$, and $z$ axes,
the index of the conformation $l\in \L$, and the scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.
We assume that the ligand can change from conformation $l_1$ at configuration $q_1$ to a conformation $l_2$ at configuration $q_2$
if both $q_1$ and $q_2$ are collision-free and if the distance from $q_1$ to $q_2$ is less than the resolution of the planner ($\varepsilon=0.05$~\AA).

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various spatial conformations.
The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
The initial configurations are searched around the beginning of the tunnel.
A random sample $q$ is generated around the first sphere of the tunnel 
$c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
If the sample $q$ is collision-free, it can be considered as a new starting configuration and it is added to $\QI$.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.

@@

The second modification is the expansion procedure (Alg.~\ref{alg::expand}). % which expands the tree from $\qnear$ towards $\qrand$.
%The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
The traditional straight-line expansion (i.e., new configuration is constructed on line from $\qnear$ to $\qrand$ in the distance
        $\varepsilon$ from $\qnear$) more likely lead to collisions, especially in narrow void space.
%Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line in the 
%distance $\varepsilon$ from $\qnear$.
%However, in the case of ligands, this would mostly lead to a collision.

The proposed expansion attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
and tested for collisions.
The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
part $(x,y,z)$ is generated around $\qnear$.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales.
This forces the planner to automatically use smaller scales in the difficult (narrow) parts of the tunnels, while preferring
large scales if possible.

@@

Each conformation from $\L$ is tested during the tree expansion. %in order to approach $\qrand$.
By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
Using a large number of conformations also increases the time complexity of the expansion step.
Selection of a suitable set of conformations $\L$ is therefore crucial.
Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are closer to each other, but 
the atoms can also form a more elongated shape (Fig.~\ref{fig::m040c}).
From the motion planning point of view, both types of conformations are useful.
The longer conformations can better fit in a narrow passage, while the packed conformations can be better rotated at a given position.
In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
packed conformations can be used.
%\vspace{-3mm}

\begin{figure}[hb]
\centering
\includegraphics[width=0.14\textwidth]{fig/m040-conf1} \hskip 25pt
\includegraphics[width=0.11\textwidth]{fig/m040-conf2}
\caption{\label{fig::m040c}
    \small
 The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
}
\end{figure}

@@

The proposed method was used to compute ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (PDB ID 1CQW) which consists of 4650 atoms.
The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
MD of the protein was computed using the AMBER tool (details about the MD simulations are described in~\cite{marques2017catalytic}).
In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
In most of the frames, two tunnels were detected.
The average bottleneck sizes of these tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the main and the side tunnel, respectively.
The tunnel traversability was analyzed in 100 consecutive frames of MD simulation in part where the protein was stabilized.

@@

Two motion planners were compared: the proposed planner (described in Alg.~\ref{alg::main} and Alg.~\ref{alg::expand}), denoted as \RA\ in the rest of the paper, and the Retraction-based RRT~\cite{zhangRetraction}, denoted as \RB\ in the rest of the paper.
The \RA\ method was run with the following parameters:
$\Imax=5,000$, $m=100$, $\gb=0.9$, $\dt=1.5$~\AA, $\rv=2$~\AA\ and $\varepsilon=0.05$~\AA.
Both methods utilize the same data structures and low-level routines (e.g., collision-detection) to enable fair comparison.
%were implemented in the same software framework and they utilized the same data structures and routines for the collision detection and nearest-neighbor search.
The nearest-neighbor search of both planners was based on the 6D Euclidean metric with the same weights for rotation and translation parts.
This metric was selected based on preliminary experiments with both planners as it resulted in the highest success rates.

@@

\begin{table}[ht]
\centering
\caption{\label{tab::one}
    \small
    Traversability for one atom.
}
\renewcommand{\tabcolsep}{4.3pt}
{\small
\renewcommand{\arraystretch}{0.7}
\input graphProbe.tex
}
\end{table}

@@

\begin{table}[ht]
\centering
\caption{\label{tab::selection}
    \small
    The influence of the conformation selection method to the traversability rates in the first tunnel.
    The number after~`$/$' denotes the number of atoms.
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\small
\renewcommand{\arraystretch}{0.7}
\input graphSel.tex
}
\end{table}

@@

Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane), and
the task was to compute the trajectories for these ligands in a given molecular dynamics simulation.
Models of the ligands in Mol2 in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand} containing the topology, rotatable bonds, atom types and partial charges. 
Rotamer library was created by rotation of all rotatable bonds with 30$^\circ$ step. 
The energy cut-off for individual conformations was set to 0.075 %(-rot_ensemble_ecutoff 0.075) 
and Rosetta database 3.2.1 was used during the conformers generation.

%\cite{meiler2006rosettaligand}
%All substrates were prepared by XXXXXX program and saved in mol2 file. Mol2 formats were converted to param files [1] containing the topology, rotatable bonds, atom types and partial charges by molfile_to_params.py script supplemented in Rosetta Scripts [2].

%Rotamer library was created by rotation of all rotatable bonds with 30$^{\circ}$ step. 
%The energy cut-off for individual conformations was set to 0.075. % (-rot_ensemble_ecutoff 0.075) 
%and Rosetta database 3.2.1 was used during the conformers generation.
%Conformations passing the energy criteria were output as multi-pdb file.

@@

\begin{table}[bt]
\centering
\caption{\label{tab::main}
    \small
    Traversability of the tunnels for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\small
\renewcommand{\arraystretch}{0.7}
\input graph.tex
}
\end{table}

@@

{\def\a{270}
\begin{figure}[hb]
\centering
%\includegraphics[width=0.17\textwidth,angle=\a]{fig/m037tmins-03tunnel-1rrt_path_time-2}
%\includegraphics[width=0.17\textwidth,angle=\a]{fig/m037tmins-03tunnel-1trajDistToGoal-2}
\includegraphics[width=0.18\textwidth,angle=\a]{fig/crop1}
\includegraphics[width=0.18\textwidth,angle=\a]{fig/crop2}
\caption{\label{fig::comparison}
    \small
    Runtimes and the distance to the active site for the m037 ligand at scale 0.6.
    The values are computed from all 200 trials (20 initial configurations $\times$ 10 trials).
    Only each third frame is shown in the graph.
}
\end{figure}
}
